Factors affecting reliablity of state-space age-structured assessment models

### 2025-03-13 Project_0/manuscript/manuscript.Rmd
saving work on project 0 paper

Application of state-space models in fisheries stock assessment and management has expanded dramatically within ICES, Canada, and the Northeast US [@nielsenberg14;@cadigan16;@pedersenberg17;@stockmiller21]. State-space approaches that use random effects to parameterize process errors is considered best practice and a requirement for the next generation of stock assessment models [@hoyleetal22; @punt23].

Although the statistical aspects of state-space models and their application have been studied extensively, the work is primarily on Gaussian state-space models. State-space stock assessment models, where there are nonlinear functions of latent parameters and multiple types of observations with varying distributional assumptions, is one of the most complex examples of this analytical approach.

Much is known about the reliability of state-space models that are linear or Gaussian [@aeberhardetal18], but not for the more complex application in stock assessment modeling.

[@lietal24 investigated some aspects of inferences for operating models with multiple sources of process error, but there are differences for this paper.]

There is a wide range of model parameters that can be treated as random effects in assessment models and a variety of potential distributions for the random effects. We know relatively little about the factors affecting statistical reliability of such models or the ability to distinguish among such alternative structures.

<!-- But those studies focus primarily on Gaussian linear models.
review literature on reliability of hidden/latent process models. Primarily in other fields.
-->

Here we conduct a simulation study with operating models (OMs) varying by degree of observation error error, source and variability of process error, and fishing history. The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated. We evaluate whether AIC can correctly determine the correct source of process error and a stock-recruit relationship. We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.

@@

Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival (R and R+S). Fitting models to other OMs rarely preferred R+S EMs and R and R+S EMs were nearly always selected for the matching OMs. For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.

@@

>>>>

Application of state-space models in fisheries stock assessment and management has expanded dramatically within International Council for the Exploration of the Sea (ICES), Canada, and the Northeast US [@nielsenberg14; @cadigan16; @pedersenberg17; @stockmiller21]. State-space models latent population characteristics as statistical time series with periodic observations that also may have error due to sampling or other sources of measurement error. Traditional assessment models may use state-space approaches to account for temporal variability in population characteristics [@methotwetzel13;@legaultrestrepo99], but these models treat the annual parameters as penalized fixed effects parameters where the variance parameters controlling the penalties are assumed known [@thorsonminto15]. Modern state-space models can estimate the annually varying parameters as random effects with variance parameters estimated using maximum marginal likelihood or corresponding Bayesian approaches. These latter approaches are considered best practice  and a requirement for the next generation of stock assessment models [@hoyleetal22; @punt23].

State-space stock assessment models, with nonlinear functions of latent parameters and multiple types of observations with varying distributional assumptions, are one of the most complex examples of this analytical approach. Statistical aspects of state-space models and their application within fisheries have been studied extensively, but previous work has focused primarily on linear and Gaussian state-space models [@aeberhardetal18; @augeretal21]. Therefore, current understanding of the reliability of state-space models does not extend to usage for stock assessment.

As state-space models provide greater flexibility by allowing multiple processes to vary as random effects [@nielsenberg14; @aeberhardetal18; @stocketal21], one of the most immediate questions regards the implications of mis-specification among alternative sources of process error. Incorrect treatment of population attributes as temporally varying [@trijouletetal20; @liljestrandetal24] could lead to misidentification of stock status and biased population estimates, ultimately impacting fisheries management decisions [@legaultpalmer16; @szuwalskietal18; @croninpunt21]. Furthermore, biological, fishery, and observational processes are often confounded in catch-at-age data, which may adversely affect ability to distinguish between true process variability and observational error [@puntetal14; @stewartmonnahan17; @croninpunt21; @fischetal23; @lietalinreview_a].

@lietal24 conducted a full-factorial simulation-estimation study to assess model reliability when confounding random-effects processes (numbers-at-age, fishery selectivity, and natural mortality) were included. Their results suggest that while state-space models can generally identify sources of process error, overly complex models, even when misspecified (i.e., incorporating process error that did not exist in reality), often performed similarly to correctly specified models, with little to no bias in key management quantities. Similarly, @liljestrandetal24 found little downside in assuming process error in recruitment or selectivity, even when it was absent.

Despite mounting efforts, several limitations remain. First, confounding processes that can be treated as random effects in the model were not thoroughly examined or tested within a simulation-estimation framework. Second, previous studies relied on operating models conditioned on specific fisheries, limiting their generalizability [@lietalinreview_a; @liljestrandetal24]. In particular, the effects of observation error and underlying fishing history were not fully isolated from their study design, making it challenging to disentangle the interplay between process and observation error magnitudes, as demonstrated in @fischetal23. Third, explicitly modeling stock-recruitment relationships as mechanistic drivers of population dynamics is promising [@fleischmanetal13; @dupontaviceetal22] but can be challenging in state-space models. Evidence suggests that when both process and observation errors are unknown, estimating density dependence parameters becomes highly uncertain [@knape08; @polanskyetal09]. In particular, @knape08 demonstrated that stronger density dependence becomes increasingly difficult to estimate in the presence of observation error. Therefore, it is crucial to assess whether density dependence mechanisms can be estimated with sufficient precision before drawing ecological conclusions [@augeretal16].

In the present study, we conduct a simulation study with operating models (OMs) varying by degree of observation error, source and variability of process error, and fishing history. The simulations from these OMs are fitted with estimation models (EMs) that make alternative assumptions for sources of process error, whether a stock-recruit model was estimated, and whether a constant, or, in some EMs, median, natural mortality is estimated. Given the confounding nature of process errors, developing diagnostic tools to detect model misspecification is of great scientific interest and could aid the next generation of stock assessments [@augeretal21]. We evaluate whether convergence and Akaike Information Criterion (AIC) can correctly determine the source of process error and the existence of a stock-recruit relationship. We also evaluate when retrospective patterns occur and the degree of bias in the outputs of the assessment model that are important for management.

@@

Among the process error configurations we used in OMs, we found AIC to be accurate for process errors on recruitment and survival (R and R+S). Fitting models to other OMs rarely preferred R+S EMs, and R and R+S EMs were nearly always selected for the matching OMs, a result similar to that reported in [@liljestrandetal24]. For other sources of process error, accuracy of AIC was improved to useful levels when there was larger variability in the process errors and/or lower observation error.

AIC performed poorly in identifying stock-recruit relationships unless there was contrast in SSB and low variability in recruitment process errors (Fig. 3), both of which help reduce confounding effects from both process and observation errors. OMs with increased  contrast in SSB included higher variability in survival or natural mortality process errors (R+S, R+M OMs), and contrast in fishing pressure (Fig. \ref{sr_aic}). Process errors in index catchability (R+q OMs) do not affect the contrast in SSB and variability in fishery logistic selectivity parameters (R+Sel OMs) had lesser effect on constrast in SSB.

Increased observation error only weakens AIC performance for OMs where process error configurations allows SR accuracy in low observation error

@@

Our findings also suggest that process errors in selectivity and catchability have minimal impact on the AIC accuracy for the stock-recruit relationship (Fig. 3C&D), whereas natural mortality process errors have a much stronger effect (Fig. 3B). This may be because selectivity and catchability do not disproportionately favorare lower for the younger age classes, and vulnerability of older ages follows an increasing logistic curve. In contrast, natural mortality influences all age classes equally, leading to greater overall fluctuations in SSB, which increases the probability of detecting the stock-recruit signal even under constant fishing pressure (Fig. 3B). These results highlight the importance of accounting for process error in natural mortality (only when there is evidence of temporal variability, [@li2024]) when evaluating stock-recruitment dynamics in state-space models.

### 2025-03-13 Project_0/manuscript/manuscript.tex
saving work on project 0 paper

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{13 February, 2025}

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within ICES, Canada, and the
Northeast US (Nielsen and Berg 2014; Cadigan 2016; Pedersen and Berg
2017; Stock and Miller 2021). State-space approaches that use random
effects to parameterize process errors is considered best practice and a
requirement for the next generation of stock assessment models (Hoyle et
al. 2022; Punt 2023).

Although the statistical aspects of state-space models and their
application have been studied extensively, the work is primarily on
Gaussian state-space models. State-space stock assessment models, where
there are nonlinear functions of latent parameters and multiple types of
observations with varying distributional assumptions, is one of the most
complex examples of this analytical approach.

Much is known about the reliability of state-space models that are
linear or Gaussian (Aeberhard et al. 2018), but not for the more complex
application in stock assessment modeling.

(Li et al. 2024 investigated some aspects of inferences for operating
models with multiple sources of process error, but there are differences
for this paper.)

There is a wide range of model parameters that can be treated as random
effects in assessment models and a variety of potential distributions
for the random effects. We know relatively little about the factors
affecting statistical reliability of such models or the ability to
distinguish among such alternative structures.

Here we conduct a simulation study with operating models (OMs) varying
by degree of observation error error, source and variability of process
error, and fishing history. The simulations from these OMs are fitted
with estimation models (EMs) that make alternative assumptions for
sources of process error, whether a stock-recruit model was estimated,
and whether a constant, or, in some EMs, median, natural mortality is
estimated. We evaluate whether AIC can correctly determine the correct
source of process error and a stock-recruit relationship. We also
evaluate when retrospective patterns occur and the degree of bias in the
outputs of the assessment model that are important for management.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

>>>>

\title{Factors affecting reliablity of state-space age-structured
assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{12 March, 2025}

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within International Council for
the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
2021). State-space models latent population characteristics as
statistical time series with periodic observations that also may have
error due to sampling or other sources of measurement error. Traditional
assessment models may use state-space approaches to account for temporal
variability in population characteristics (Legault and Restrepo 1999;
Methot and Wetzel 2013), but these models treat the annual parameters as
penalized fixed effects parameters where the variance parameters
controlling the penalties are assumed known (Thorson and Minto 2015).
Modern state-space models can estimate the annually varying parameters
as random effects with variance parameters estimated using maximum
marginal likelihood or corresponding Bayesian approaches. These latter
approaches are considered best practice and a requirement for the next
generation of stock assessment models (Hoyle et al. 2022; Punt 2023).

State-space stock assessment models, with nonlinear functions of latent
parameters and multiple types of observations with varying
distributional assumptions, are one of the most complex examples of this
analytical approach. Statistical aspects of state-space models and their
application within fisheries have been studied extensively, but previous
work has focused primarily on linear and Gaussian state-space models
(Aeberhard et al. 2018; Auger-Méthé et al. 2021). Therefore, current
understanding of the reliability of state-space models does not extend
to usage for stock assessment.

As state-space models provide greater flexibility by allowing multiple
processes to vary as random effects (Nielsen and Berg 2014; Aeberhard et
al. 2018; Stock et al. 2021), one of the most immediate questions
regards the implications of mis-specification among alternative sources
of process error. Incorrect treatment of population attributes as
temporally varying (Trijoulet et al. 2020; Liljestrand et al. 2024)
could lead to misidentification of stock status and biased population
estimates, ultimately impacting fisheries management decisions (Legault
and Palmer 2016; Szuwalski et al. 2018; Cronin-Fine and Punt 2021).
Furthermore, biological, fishery, and observational processes are often
confounded in catch-at-age data, which may adversely affect ability to
distinguish between true process variability and observational error (Li
et al. In review; Punt et al. 2014; Stewart and Monnahan 2017;
Cronin-Fine and Punt 2021; Fisch et al. 2023).

Li et al. (2024) conducted a full-factorial simulation-estimation study
to assess model reliability when confounding random-effects processes
(numbers-at-age, fishery selectivity, and natural mortality) were
included. Their results suggest that while state-space models can
generally identify sources of process error, overly complex models, even
when misspecified (i.e., incorporating process error that did not exist
in reality), often performed similarly to correctly specified models,
with little to no bias in key management quantities. Similarly,
Liljestrand et al. (2024) found little downside in assuming process
error in recruitment or selectivity, even when it was absent.

Despite mounting efforts, several limitations remain. First, confounding
processes that can be treated as random effects in the model were not
thoroughly examined or tested within a simulation-estimation framework.
Second, previous studies relied on operating models conditioned on
specific fisheries, limiting their generalizability (Li et al. In
review; Liljestrand et al. 2024). In particular, the effects of
observation error and underlying fishing history were not fully isolated
from their study design, making it challenging to disentangle the
interplay between process and observation error magnitudes, as
demonstrated in Fisch et al. (2023). Third, explicitly modeling
stock-recruitment relationships as mechanistic drivers of population
dynamics is promising (Fleischman et al. 2013; Du Pontavice et al. 2022)
but can be challenging in state-space models. Evidence suggests that
when both process and observation errors are unknown, estimating density
dependence parameters becomes highly uncertain (Knape 2008; Polansky et
al. 2009). In particular, Knape (2008) demonstrated that stronger
density dependence becomes increasingly difficult to estimate in the
presence of observation error. Therefore, it is crucial to assess
whether density dependence mechanisms can be estimated with sufficient
precision before drawing ecological conclusions (Auger-Méthé et al.
2016).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a stock-recruit model
was estimated, and whether a constant, or, in some EMs, median, natural
mortality is estimated. Given the confounding nature of process errors,
developing diagnostic tools to detect model misspecification is of great
scientific interest and could aid the next generation of stock
assessments (Auger-Méthé et al. 2021). We evaluate whether convergence
and Akaike Information Criterion (AIC) can correctly determine the
source of process error and the existence of a stock-recruit
relationship. We also evaluate when retrospective patterns occur and the
degree of bias in the outputs of the assessment model that are important
for management.

@@

\leavevmode\vadjust pre{\hypertarget{ref-augeretal16}{}}%
Auger-Méthé, M., Field, C., Albertsen, C.M., Derocher, A.E., Lewis,
M.A., Jonsen, I.D., and Mills Flemming, J. 2016. State-space models'
dirty little secrets: Even simple linear gaussian models can have
estimation problems. Scientific reports \textbf{6}(1): 26677.
doi:\href{https://doi.org/10.1038/srep26677}{10.1038/srep26677}.

\leavevmode\vadjust pre{\hypertarget{ref-augeretal21}{}}%
Auger-Méthé, M., Newman, K., Cole, D., Empacher, F., Gryba, R., King,
A.A., Leos-Barajas, V., Mills Flemming, J., Nielsen, A., Petris, G., and
others. 2021. A guide to state--space modeling of ecological time
series. Ecological Monographs \textbf{91}(4): e01470.
doi:\href{https://doi.org/10.1002/ecm.1470}{10.1002/ecm.1470}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-croninpunt21}{}}%
Cronin-Fine, L., and Punt, A.E. 2021. Modeling time-varying selectivity
in size-structured assessment models. Fisheries Research \textbf{239}:
105927. Elsevier.

@@

\leavevmode\vadjust pre{\hypertarget{ref-dupontaviceetal22}{}}%
Du Pontavice, H., Miller, T.J., Stock, B.C., Chen, Z., and Saba, V.S.
2022. Ocean model-based covariates improve a marine fish stock
assessment when observations are limited. ICES Journal of Marine Science
\textbf{79}(4): 1259--1273.
doi:\href{https://doi.org/10.1093/icesjms/fsac050}{10.1093/icesjms/fsac050}.

\leavevmode\vadjust pre{\hypertarget{ref-fischetal23}{}}%
Fisch, N., Shertzer, K., Camp, E., Maunder, M., and Ahrens, R. 2023.
Process and sampling variance within fisheries stock assessment models:
Estimability, likelihood choice, and the consequences of incorrect
specification. ICES Journal of Marine Science \textbf{80}(8):
2125--2149. Oxford University Press.

\leavevmode\vadjust pre{\hypertarget{ref-fleischmanetal13}{}}%
Fleischman, S.J., Catalano, M.J., Clark, R.A., and Bernard, D.R. 2013.
An age-structured state-space stock--recruit model for pacific salmon
(oncorhynchus spp.). Canadian Journal of Fisheries and Aquatic Sciences
\textbf{70}(3): 401--414. NRC Research Press.

@@

\leavevmode\vadjust pre{\hypertarget{ref-knape08}{}}%
Knape, J. 2008. Estimability of density dependence in models of time
series data. Ecology \textbf{89}(11): 2994--3000.
doi:\href{https://doi.org/10.1890/08-0071.1}{10.1890/08-0071.1}.

\leavevmode\vadjust pre{\hypertarget{ref-legaultpalmer16}{}}%
Legault, C.M., and Palmer, M.C. 2016. In what direction should the
fishing mortality target change when natural mortality increases within
an assessment? Canadian Journal of Fisheries and Aquatic Sciences
\textbf{73}(3): 349--357. NRC Research Press.

\leavevmode\vadjust pre{\hypertarget{ref-legaultrestrepo99}{}}%
Legault, C.M., and Restrepo, V.R. 1999. A flexible forward
age-structured assessment program. Col. Vol. Sci. Pap. ICCAT
\textbf{49}(2): 246--253.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lietalinreview_a}{}}%
Li, C., Deroba, J.J., Berger, A.M., Goethel, D.R., Langseth, B.J.,
Schueller, A.M., and Miller, T.J. In review. Random effects on
numbers-at-age transitions implictly 2 account for movement dynamics and
improve 3 performance within a state-space stock assessment. Canadian
Journal of Fisheries and Aquatic Sciences.

@@

\leavevmode\vadjust pre{\hypertarget{ref-liljestrandetal24}{}}%
Liljestrand, E.M., Bence, J.R., and Deroba, J.J. 2024. The effect of
process variability and data quality on performance of a state-space
stock assessment model. Fisheries Research \textbf{275}: 107023.
Elsevier.

@@

\leavevmode\vadjust pre{\hypertarget{ref-methotwetzel13}{}}%
Methot, R.D., and Wetzel, C.R. 2013. Stock synthesis: A biological and
statistical framework for fish stock assessment and fishery management.
Fisheries Research \textbf{142}: 86--99.
doi:\href{https://doi.org/10.1016/j.fishres.2012.10.012}{10.1016/j.fishres.2012.10.012}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-polanskyetal09}{}}%
Polansky, L., De Valpine, P., Lloyd-Smith, J.O., and Getz, W.M. 2009.
Likelihood ridges and multimodality in population growth rate models.
Ecology \textbf{90}(8): 2313--2320.
doi:\href{https://doi.org/10.1890/08-1461.1}{10.1890/08-1461.1}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-puntetal14}{}}%
Punt, A.E., Hurtado-Ferro, F., and Whitten, A.R. 2014. Model selection
for selectivity in fisheries stock assessments. Fisheries Research
\textbf{158}: 124--134. Elsevier.

@@

\leavevmode\vadjust pre{\hypertarget{ref-stewartmonnahan17}{}}%
Stewart, I.J., and Monnahan, C.C. 2017. Implications of process error in
selectivity for approaches to weighting compositional data in fisheries
stock assessments. Fisheries Research \textbf{192}: 126--134. Elsevier.

@@

\leavevmode\vadjust pre{\hypertarget{ref-stocketal21}{}}%
Stock, B.C., Xu, H., Miller, T.J., Thorson, J.T., and Nye, J.A. 2021.
{Implementing two-dimensional autocorrelation in either survival or
natural mortality improves a state-space assessment model for Southern
New England-Mid Atlantic yellowtail flounder}. Fisheries Research
\textbf{237}: 105873.

\leavevmode\vadjust pre{\hypertarget{ref-szuwalskietal18}{}}%
Szuwalski, C.S., Ianelli, J.N., and Punt, A.E. 2018. Reducing
retrospective patterns in stock assessment and impacts on management
performance. ICES Journal of Marine Science \textbf{75}(2): 596--609.
Oxford University Press.

\leavevmode\vadjust pre{\hypertarget{ref-thorsonminto15}{}}%
Thorson, J.T., and Minto, C. 2015. Mixed effects: A unifying framework
for statistical modelling in fisheries biology. ICES Journal of Marine
Science \textbf{72}(5): 1245--1256.
doi:\href{https://doi.org/10.1093/icesjms/fsu213}{10.1093/icesjms/fsu213}.

\leavevmode\vadjust pre{\hypertarget{ref-trijouletetal20}{}}%
Trijoulet, V., Fay, G., and Miller, T.J. 2020. Performance of a
state-space multispecies model: What are the consequences of ignoring
predation and process errors in stock assessments? Journal of Applied
Ecology \textbf{57}(1): 121--135. Wiley Online Library.
